---
title: "A pedigree-transmission likelihood for multiplex families"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  pdf_notebook: default
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(warning=FALSE, message=FALSE)
```

This document presents a tutorial describing Tianyu's work in summer 2022 to implement a pedigree-transmission likelihood. We are interested in transmission-parameter
likelihoods to evaluate the evidence that a rare variant (RV) is causally linked to the disease in an extended family, or pedigree. To implement the likelihood we use  the `gRain` R package [1] to propagate probability in Bayesian networks. Bayesian networks are a special case of $gRa$phical $i$ndependence $n$etworks. In graphical independence networks, 
random variables are nodes in a graph 
and conditional dependencies are edges between nodes,
allowing a compact representation of 
the joint probability distribution of the random variables.
The graphical independence networks in the gRain package are restricted to comprising 
discrete variables, each with a finite state space.

The project's goal is to calculate the joint 
probability of an RV configuration in the diseased individuals with DNA available in the pedigree.
A pedigree can be viewed as a **directed** graphical 
model of the RV status of its members, with the RV
status of a child depending causally 
on the RV status of its
parents. Thus a pedigree can be 
cast as a Bayesian network,
a directed acyclic graph.
We use `gRain` because it allows us to cast a pedigree as a Bayesian network
and more easily calculate transmission-parameter likelihoods. 

To start, we load the required packages: `gRain` and the pedigree drawing and manipulation R package `kinship2` [2].
```{r}
library(gRain)
library(kinship2)
```


# 1. Getting to know the `gRain` package

We work through a simple example 
showing how to use `gRain` functions to calculate the 
probability that the affected individuals in a pedigree
received a rare variant (RV) from a carrier founder, 
given that the RV is transmitted with probability 3/4. 

The example pedigree is shown below, drawn with
the `pedigree()` function of the `kinship2` package.
The first step for plotting is to label each family 
member with an ID number and indicate their father and mother.
```{r}
id = c(1, 2, 3, 4, 5, 6, 7, 8, 9)
# Label each family member with id number
father = c(0, 0, 1, 0, 1, 0, 3, 5, 5)
# Use id number to indicate the father of each family member
mother = c(0, 0, 2, 0, 2, 0, 4, 6, 6)
# Use id number to indicate the mother of each family member
```

Next indicate the biological sex, affection status 
and DNA availability of each individual. In our applications,
DNA is typically available only from living disease-affected members
of the pedigree. The living disease-affected members 
tend to be in the more recent generations of a pedigree, and are typically non-founders. 
```{r}
sex = c(1, 2, 1, 2, 1, 2, 2, 2, 2)
# 1 represents male, 2 represents female.
affected = c(0, 0, 0, 0, 0, 0, 1, 1, 1) 
# affected specifies which family members are affected by the disease.
# 0 and 1 represent individuals who don't and do have the disease, respectively.
avail = c(0, 0, 0, 0, 0, 0, 1, 1, 1) 
# avail specifies which family members have DNA available.
# 0 and 1 represent individuals who don't and do have DNA available, respectively.
```

Assemble the information into a data frame for the pedigree and then convert
the data frame into a pedigree object.
```{r}
toyPed = as.data.frame(cbind(id,father,mother,sex,avail, affected))

toyPed_ob = pedigree(toyPed$id, toyPed$father, toyPed$mother, toyPed$sex, 
                     affected=cbind(toyPed$affected, toyPed$avail))
# Use pedigree() function from the kinship2 package to construct the pedigree object. 
```

Finally, plot the resulting pedigree with affection status marked on the left and DNA availability marked on the right of nodes for individuals.
```{r}
plot(toyPed_ob)
```

From the output above, we can see that ID's 7, 8 and 9 are the only affected members and the only ones with DNA available. 

We would like to cast this pedigree as a Bayesian network.
A Bayesian network models the joint distribution of a set of variables in terms of their conditional dependencies 
in a directed graph. A network consists of nodes (variables)
and components known as conditional probability tables (CPTs). CPTs specify the local conditional dependencies
between variables in the network. 
When the dependencies among variables are
local in nature, it is a lot easier to understand
them in terms of CPTs than 
in terms of the joint distributions of all the variables. 
 
In our context, the variables are the RV status of 
individuals in a pedigree and the pedigree is the directed
graph. The conditional probability tables specify 
the local dependency of a child's RV genotype given
the RV genotypes of their two parents. For pedigrees 
with no inbreeding, 
the local dependency structure is straightforward,
with children depending only on their parents.

To set up the Bayesian network, we first specify
the levels for the RV status of pedigree members
as 0, 1 or 2 copies of the RV.

```{r}
copy = c("0", "1", "2")
# number of variants levels: 0, 1, 2
```

Next we consider all $3^3=27$ 
possible configurations of the RV status
for the child and its two parents. For each
configuration, we specify the conditional probability of the child's
RV status given the RV status of the two parents assuming
that the transmission probability of the RV from a parent
to the child is $\tau=3/4$. We will use these conditional probabilities
to set up our CPTs.
```{r}
# conditional probability of child in the first entry of the 
# triplet marked below, given parents in the second and third 
# entries of triplet.
child_1 = c(1, 0, 0, 1/4, 3/4, 0, 0, 3/4, 0)
# 000 100 200 010 110 210 020 120 220 
child_2 = c(1/4, 3/4, 0, 1/16, 3/8, 9/16, 0, 1/4, 3/4)
# 001 101 201 011 111 211 021 121 221 
child_3 = c(0, 1, 0, 0, 1/4, 3/4, 0, 0, 1)
# 002 102 202 012 112 212 022 122 222
child = c(child_1, child_2, child_3)
```

We may now specify the CPTs with the values given in the `child` vector. We start with the pedigree non-founders. These
are the individuals who have information on both parents.
According to the pedigree diagram above, the non-founders are 
IDs 3, 5, 7, 8 and 9. The non-founders with IDs 3 and 5
are siblings and have parents with IDs 1 and 2.
The non-founder with ID 7 has parents with IDs 3 and 4.
Finally, the non-founders with IDs 8 and 9 are siblings and have
parents with IDs 5 and 6. In the CPT `formula` argument, the child node comes first after the `~`, followed by the parent nodes,
because the child's RV status is conditionally 
dependent on the parents' RV status.

```{r}
# CPTs for non-founders
ID_3.12 = cptable(~3 + 1 + 2, values = child, levels = copy)
ID_5.12 = cptable(~5 + 1 + 2, values = child, levels = copy)
ID_7.34 = cptable(~7 + 3 + 4, values = child, levels = copy)
ID_8.56 = cptable(~8 + 5 + 6, values = child, levels = copy)
ID_9.56 = cptable(~9 + 5 + 6, values = child, levels = copy)
# Note the order of the variables after the + operator matters.
# The child nodes come first after ~ followed by the parents.
```

We then set up the conditional probability tables for the
founders of the pedigree. 
According to the pedigree diagram above, the 
founders are 
IDs 1, 2, 4 and 6.
Since founders have no parents, we specify the
marginal probability of their RV genotypes in their CPT 
tables. Our application does not require
the marginal RV probabilities for the founders 
but the `gRain` package does.
We therefore set the three RV 
genotypes in founders to be equiprobable. 

```{r}
# Conditional probability tables for founders.
ID_1 = cptable(~1, values = rep(1/3, 3), levels = copy)
ID_2 = cptable(~2, values = rep(1/3, 3), levels = copy)
ID_4 = cptable(~4, values = rep(1/3, 3), levels = copy)
ID_6 = cptable(~6, values = rep(1/3, 3), levels = copy)
# We don't know the values for founders since they don't 
# have parents in our pedigree. 
```

Next we combine all the CPTs into a list for our Bayesian network:
```{r}
plist_pedigree <- 
    compileCPT(list(ID_1, ID_2, ID_3.12, ID_4, ID_5.12, 
                    ID_6, ID_7.34, ID_8.56, ID_9.56))
```

From this combined CPT list we can extract the individual
CPTs of individuals in the pedigree. For example, we can 
extract the CPT for the individual with ID 3 as follows. 
```{r}
plist_pedigree$`3`
```
The output gives the conditional probabilities that ID 3 
has 0, 1, or 2 copies of the RV, given the RV genotype
status of its parents ID 1 and ID 2.

We are ready to create the network from the list of CPTs.
```{r}
gin1 = grain(plist_pedigree)
summary(gin1)
```
The `summary()` of the network tells us it has
five "cliques", corresponding to the CPT parent-child trios for the five children
in the pedigree. The maximal state-space in these cliques
is $3^3=27$, corresponding to possible RV-status
outcomes for the child-parent trios.

We now have a Bayesian network for efficiently
calculating joint RV probabilities. Being rare, the RV is assumed to enter the pedigree through a single founder. For illustration, let's assume the founder who introduced the RV into the pedigree is ID 1. 
The next code chunk calculates the conditional joint probability
that all three of the affected individuals carry the RV
given that the founder with ID 1 introduced it into the pedigree. The RV transmission probability is taken
to be $\tau=3/4$. 
We focus on the affected individuals because in medical-genetic studies 
they are often the only pedigree members who are genotyped
for RV status. 

```{r}
#Condition on founder id1 introducing the RV into the pedigree.
bn1 = setEvidence(gin1, nodes = c("1", "2", "4", "6"), 
                  states = c("1", "0", "0", "0") )
```
In Bayesian networks, "evidence" is what is being conditioned on to get the probability. We use the 
`setEvidence()` function to condition on ID 1
being the founder who introduced the single copy of
the RV into the pedigree. The other founders IDs 2, 4, and 6 introduce no copies of the 
RV into the pedigree.

Let's start by getting the marginal
distributions of RV status for each of the three affected 
individuals.

```{r}
affs=c("7", "8", "9")
querygrain(bn1, nodes = affs, type = "marginal")
```

Next, we move on to get the joint probability distribution for 
the three affected individuals.
```{r}
config=c("1","1", "1") #RV configuration for affected members
names(config)=affs #link config to the affected IDs  
querygrain(bn1, nodes=affs, type="joint")
```
The output lists the conditional joint distribution 
of the RV status of the affected individuals given
that the founder with ID 1 introduced the
RV into the pedigree. This probability is calculated
assuming that the transmission
probability of the RV is $\tau=3/4$. We see that
the probability that all three of the affected individuals
carry a single copy of the RV given that the RV was introduced
into the pedigree by the founder with ID 1 is 0.2373. We may
also get this result by repeated conditioning, also known as the "chain rule" for Bayesian networks.

```{r}
p = unlist(querygrain(bn1, nodes = "7", type = "marginal"))
names(p) = copy 
p[config["7"]]
```

```{r}
prob=1
for(n in affs){
# calculate the marginal probability for the node, n 
p=unlist(querygrain(bn1, nodes=n))
names(p)=copy
prob=prob*p[config[n]]
# condition on node n's state
bn1 = setEvidence(bn1, nodes=n, states=config[n])
}
prob
```
The output confirms that the 
probability that all three of the affected individuals
carry a single copy of the RV introduced 
by ID 1 is 0.2373.


# 2. Function to create a Bayesian network

In this section, we present a function to create a Bayesian network to be used in our application. The function takes two arguments:

1. A `kinship2` pedigree dataframe, `pedfile`, and
2. A transmission probability of the RV, `tau`.

This function `BNcreate()` is defined in
the code chunk below.
```{r}
BNcreate = function(pedfile, tau){
  ## conditional prob of child in the first entry of the triplet 
  ## given parents in the second and third entries of triplet with transmission 
  ## probability tau
  geno_prob = c(
    # 000 100 200 010 110 210 020 120 220
    1, 0, 0, 1-tau, tau, 0, 0, 1, 0,
    # 001 101 201 011 111 211 021 121 221
    1-tau, tau, 0, (1-tau)^2, 2*tau*(1-tau), tau^2, 0, 1-tau, tau,
    # 002 102 202 012 112 212 022 122 222
    0, 1, 0, 0, 1-tau, tau, 0, 0, 1
  )
   
   # Construct the CPTs for founders
   # Founders: columns of father or mother are labelled as 0.
  
   founders = which(pedfile[, "father"] == 0) # the row numbers of founders
   founders_vec = rep(0, length(founders)) 
   # store the id number of founders (for user configuration)
   founders_cpt = list() 
   # store the CPTs of founders in a list, which is denoted as [[i]]
   for (i in 1:length(founders)) {
     id = pedfile[founders[i], "id"]
     founders_vec[i] = id
     node = cptable(c(id), values = rep(1/3, 3), levels = copy) 
     # CPTs of founders
     founders_cpt[i] = list(node)
   }
  
   # Construct the CPTs for non-founders
   
   pedfile_c = pedfile[-c(founders), ]
   # Pedigree data after removing founders 
   nonfounders_cpt = list()
   # Store the CPTs for non-founders
   for (i in 1:nrow(pedfile_c)) {
     c = pedfile_c[i, 'id'] # id numbers of child
     f = pedfile_c[i, "father"] # id numbers of father
     m = pedfile_c[i, "mother"] # id numbers of mother
     node_nf = cptable(c(c, f, m), values = geno_prob, levels = copy) 
     # CPTs for non-founders
     nonfounders_cpt[i] = list(node_nf)
   }
   
   plist = compileCPT(founders_cpt, nonfounders_cpt)
   gin = grain(plist)
   # Create the Bayesian network from the CPTs of both founders and non-founders

   return(gin)
}
```


We apply the function to the pedigree dataframe from the previous example.
```{r}
gin = BNcreate(toyPed, 3/4) 
# Create the Bayesian network from the toy pedigree dataframe using a transmission probability of 3/4.
```

Let's check that we get the same summary as in the previous part
of this document, when 
we created the Bayesian network for the toy pedigree interactively.
```{r}
summary(gin) #gives same summary as above
```



# 3. Construct a function for the transmission likelihood

This next section describes how to calculate the likelihood
of a transmission probability in a multiplex 
pedigree, given data on the RV status of the 
affected members with available DNA and
assuming that a single founder introduced the
RV into the pedigree. A multiplex pedigree
is a pedigree that has been sampled or "ascertained" 
to contain several
members (e.g. 3 or more) who
are affected with the disease of interest.
Many family studies in medical genetics rely on this sampling 
design.

Our goal is to calculate the likelihood of a value
of the RV-transmission probability. For example, 
above we used a value of $\tau=3/4$ for the transmission
probability in the toy pedigree. 
As before, we assume that, being rare,
the RV enters the pedigree through a single founder only.
Therefore, in a pedigree with no inbreeding, the possible values for RV status are 0 (no copies) and 1 (a single copy).
We take the prior probability that any founder is the one that 
introduced the RV to be uniform over all the pedigree founders. 
The likelihood is obtained from 
the conditional probability of the RV configuration
for affected pedigree members with available DNA,
given that exactly one of the founders introduced
a single copy of the RV into the pedigree.

The event that exactly one founder introduced the RV is a union of the mutually exclusive events, $F_i$, that each of the founders, 
$i$, introduced the RV. These events are mutually exclusive because the chance of more than one founder carrying a RV is essentially zero. The probability of
the RV configuration $C$ in the affected pedigree members with 
available DNA is thus 
\begin{eqnarray*}
P(C | \cup_i \! F_i) & =&  \frac{P(C, \cup_i F_i)}{P(\cup_i F_i)} \\
&=& \frac{\sum_i P(C, F_i)}{\sum_i P(F_i)}\\
&=& \sum_i P(C|F_i) \frac{P(F_i)}{\sum_j P(F_j)}.
\end{eqnarray*}
Founders are assumed to have the same prior probability of introducing the RV; i.e., $P(F_1)=P(F_2)=P(F_3)=\cdots$. Thus  $P(F_i)/\sum_j P(F_j)$ on the right-hand side of the above equation is equal to one over the number of founders. The conditional
probability, $P(C|F_i)$, of the RV configuration given that founder $i$ introduced the RV is obtained by "path counting" the number 
of independent transmissions connecting founder $i$ to the affected
members. For example, in the toy pedigree above, an RV from the founder with ID 1 can reach the affected pedigree member with ID 7
through two independent transmissions from ID 1 to ID 3 and then 
from ID 3 to ID 7, and can reach IDs 8 and 9 through one independent
transmission to ID 5 and then two independent transmissions
from ID 5 to IDs 8 and 9. Therefore, $P(C=(1,1,1)| F_1) = \tau^{2+1+2}$ and $P(C=(1,1,0)|F_1) = \tau^{2+1+1}(1-\tau)$,
for example. In the calculation for $P(C=(1,1,0)|F_1)$, we
count two transmissions of the RV from ID 1 to ID 7, one
from ID 1 to ID 5, one from ID 5 to ID 8, and
one transmission of the other variant in ID 5 
(i.e. the non-RV) to ID 9.

The first step in the likelihood function is to call `BNcreate()`
to set up a Bayesian network for 
a given value of the transmission probability. 
The pedigree-transmission likelihood function takes 3 arguments:
`pedfile`, `tau` and `config` described below.
The first argument,`pedfile`,
is a `kinship2` pedigree object specifying the pedigree structure.
The second argument, `tau`,  is the transmission probability of the RV.
The third argument, `config`, is a numeric vector giving the RV status of affected individuals with DNA available. 
Each element of the
vector corresponds to an affected pedigree member with DNA. The elements are ordered by their ID numbering. For example, if the affected individuals with available DNA 
have ID numbers 1, 3, 6, 7, 14 and `config =c(1, 0, 1, 1, 1)`, 
IDs 1, 6, 7, 14 carry one copy of the RV and ID 3 carries no copies. 

We loop through all the founders, and for each
founder $i$ we calculate $P(\mbox{ config }  | \mbox { founder } i \mbox{ introduced the RV})$,
the conditional probability 
of the RV configuration given that the RV enters the pedigree on founder $i$. In the loop, we sum out over
the product of
this conditional probability and the probability that
founder $i$ is the founder who introduced the RV
to get the desired likelihood from $P(\mbox{ config }  | \mbox { exactly one founder introduced the RV})$.
```{r}
likehd = function(pedfile, tau, config){
  
  gin = BNcreate(pedfile, tau)
  # Bayesian networks of the specified pedigree with transmission prob tau
  likelihood = 0
  
  founders = which(pedfile[, "father"] == 0) 
  # find the row numbers of founders
  founders = as.character(pedfile$id[founders])
  # find the ID numbers of founders
  affs = which(pedfile[, "affected"] == 1&pedfile[,"avail"]==1) 
  # find the row numbers of affected individuals with DNA available
  affs=as.character(pedfile$id[affs])
  # find the ID numbers of affected individuals with DNA available
  
  # Assume the RV enters pedigree on exactly one founder and
  # loop through all the founders 
  for (i in 1:length(founders)) {
    state = rep(0, length(founders)) 
    # vector state specifies which founder introduces the RV 
    state[i] = "1" #1 means carrying the RV.

    bn = setEvidence(gin, nodes = founders, states = state) 
    # Set founder i to introduce the rare variant
    
    # calculate P(config | founder i introduced the rv)
    prob=1
    for(n in affs){
      if(prob > 0) # prevents conditioning on zero prob events
      { 
        # calculate probability for this node, n   
        p=unlist(querygrain(bn,nodes=n,exclude=FALSE)) #get marginal prob
        names(p)=c("0","1","2")
        prob=prob*p[config[n]]
      # condition on node n's state
        bn = setEvidence(bn, nodes=n, states=config[n])
      }
    }
    likelihood = prob*1/length(founders)+likelihood 
  }
  
  return(likelihood)
}

```

# 4. Testing the likelihood function

## 4.1 Toy pedigree

We start by testing the likelihood function
on the toy pedigree above. We evaluate
the likelihood of $\tau=3/4$ when all three
of the affected pedigree
members with available DNA carry a single
copy of the RV.
```{r}
config=c("1","1","1")
names(config)=c("7","8","9")
likehd(toyPed, 3/4, config)
```
From the output above, we see that, when 
the RV configuration is $C=(1,1,1)$, the likelihood of $\tau = 3/4$ is about $.1187$.
This value of the likelihood complies with our hand calculation from
the "path counting" approach:
\begin{eqnarray*}
P(C=(1,1,1) | \cup_i F_i)
&=& \frac{1}{4} \left \{  [\tau^2 \times \tau (\tau^2)] \; + 
\; [\tau^2 \times \tau (\tau^2)] \; + 
\; [ \tau \times 0 ] \; + 
\; [ \tau^2 \times 0 ] \right \} \\
& = & \frac{1}{4} \left \{ [\tau^5] \; + \; [\tau^5] \; + \; [0] \; + \; [0] \right \}\\
&=& \frac{1}{4}\left \{ 2 \times \left ( \frac{3}{4} \right )^5 \right \}\\
&=&.1187 
\end{eqnarray*}

Next, we evaluate the likelihood of
$\tau=3/4$ when the RV configuration for
IDs 7, 8, and 9 is $C=(1,1,0)$. In this configuration,
IDs 7 and 8 both carry the RV but ID 9 does not.

```{r}
config=c("1","1","0")
names(config)=c("7","8","9")
likehd(toyPed, 3/4, config)
```
This value of the likelihood complies with our hand calculation of $\frac{1}{2}\tau^4 (1-\tau)=\frac{1}{2} \left ( \frac{3}{4} \right )^4 \left ( 1 - \frac{3}{4} \right )=0.0396$ from the path-counting approach.

## 4.2 Modified `kinship2` pedigrees

Consider a modified version
of the second of the two 
sample pedigrees in the `kinship2` R package.
```{r}
# load kinship2 pedigree data  
data("sample.ped") #two pedigrees
testPed=sample.ped[sample.ped$ped==2,] #2nd pedigree
# modify the kinship2 pedigree data
testPed[testPed$id==203,"affected"]=0
testPed[testPed$id==211,"affected"]=1
testPed[testPed$id==202,"affected"]=0
testPed[testPed$id==201,"avail"]=0
testPed[testPed$id==203,"avail"]=0
testPed[testPed$id==204,"avail"]=0
testPed[testPed$id==212,"avail"]=0
testPed # Unknown affection status has NA
```

From the above data-frame, we see that ID
205 has unknown affection status. 
This person is marked as "?" in the 
pedigree plot below.

```{r}
testPed_ob=pedigree(testPed$id, testPed$father,testPed$mother, testPed$sex, affected=cbind(testPed$affected,testPed$avail))
plot(testPed_ob)
```

The affected family members with DNA available have IDs 206, 207, 211 and 214.
Let's evaluate the likelihood of $\tau=1/2$
for various configurations of the RV status
in these individuals. First,
suppose that all four of them share 
a copy of the RV; i.e. $C=(1,1,1,1)$.

```{r}
config1 = c("1", "1", "1", "1")
names(config1)=c("206","207","211", "214")
likehd(testPed, 1/2, config1)
```
Similarly, from the output above, we observe that, when the RV configuration is 
$C = (1,1,1,1)$, the likelihood of $\tau = 1/2$ is about $.0078125$. This likelihood
value also complies with our hand calculation from the "path counting" approach:
\begin{eqnarray*}
P(C=(1,1,1,1) | \cup_i F_i) 
&=& \frac{1}{4} \left \{  [ \tau^2 \times \tau \times \tau \times \tau^2] \; +  
\; [ \tau^2 \times \tau \times \tau \times \tau^2] \; + 
\; [1 \times \tau \times 0 ] \; +
\; [1 \times 0] \right \} \\
&=& \frac{1}{4} \left \{  [\tau^6] \; + \; [\tau^6] \; + \; [0] \; + \; [0] \right \} \\
&=& \frac{1}{4} \left \{  2 \times \left( \frac{1}{2} \right )^6 \right \} \\
&=& .0078125
\end{eqnarray*}

Next, we evaluate the likelihood of $\tau = \frac{1}{2}$ when the RV configuration 
for IDs 206, 207, 211, 214 is $C = (1,0,0,0)$, which indicates that only ID 206
carries the RV.
```{r}
config4 = c("1", "0", "0", "0")
names(config4)=c("206","207","211", "214")
likehd(testPed, 1/2, config4)
```
From the output above, we see that when the RV configuration is C = (1,0,0,0), 
the likelihood of $\tau = 1/2$ is about $.0703125$. The likelihood value also complies 
with our hand calculation from the "path counting" approach: 
\begin{eqnarray*}
P(C = (1,0,0,0) | \cup_i F_i) 
&=& \frac{1}{4} \left \{  [ \tau \times \left ( 1 - \tau \right ) \times 
\left ( \left (1 - \tau \right ) + \tau \times \left( 1 - \tau \right ) \right )^2] \;\;\;\; + \right. \\
& & \;\;\;\;\;\;\;\; \left.
\; [ \tau \times \left ( 1 - \tau \right ) \times 
\left ( \left (1 - \tau \right ) + \tau \times \left( 1 - \tau \right ) \right )^2] \; +
\; [0] \; + \; [0] \right \} \\
&=& \frac{1}{4} \left \{ \frac{1}{2} \times \frac{1}{2} \times 
\left ( \frac{1}{2} + \frac{1}{2} \times \frac{1}{2} \right )^2  \; + 
\; \frac{1}{2} \times \frac{1}{2} \times 
\left(\frac{1}{2} + \frac{1}{2} \times \frac{1}{2}\right)^2 \right \} \\
&=& .0703125
\end{eqnarray*}

Next consider `testPed2`, a version of the pedigree above in which the founder with ID 201 has DNA available. 
```{r}
testPed2=testPed
testPed2[testPed2$id==201,"avail"]=1 #Change ID 201
testPed2
```
From the output above, we see that the disease-affected founder with ID 201 now has DNA available. 

Let's work through the likelihood calculation when all
the affected pedigree members with DNA in `testPed2`
carry the RV. In this case, the RV configuration is 
$C=(1,1,1,1,1)$ for IDs 201, 206, 207, 211 and 214. 
This pedigree has four founders with IDs 201, 202, 203
and 209.
If ID 201 is the founder introducing the RV, the 
conditional probability of the configuration is
$$P(C = (1,1,1,1,1) \, | F_{201}) = 1\times \tau \times 
\tau \times \tau^2 \times \tau^2 = \tau^6.$$
If ID 202 is the founder introducing the RV, 
the conditional probability of the configuration is 
$$P(C = (1,1,1,1,1) \, | F_{202}) = 0 \times \tau \times 
\tau \times \tau^2 \times \tau^2 = 0.$$ 
Similarly, if IDs 203 or 209 are the founders
introducing
the RV, the conditional probability of the 
configuration is zero. Therefore, the likelihood 
function for the transmission probability 
is $P(C = (1,1,1,1,1) \, | \cup_i \!{F_i}) =
\frac{1}{4} \tau^6.$ This 
likelihood evaluates to $0.00390625$ for $\tau=1/2$,
as confirmed by the code chunk below:
```{r}
config5 = c("1", "1", "1", "1", "1")
names(config5)=c("201","206","207","211", "214")
likehd(testPed2, 1/2, config5)
```

Now suppose the RV configuration is $C=(0,1,1,1,1)$.
If ID 201 is the founder introducing the RV, the 
conditional probability of the configuration
is $P(C = (0,1,1,1,1) \, | F_{201})=0$ because
ID 201 not carrying the RV 
conflicts with the conditioning event that
ID 201 introduced the RV into the pedigree.
If ID 202 is the founder introducing the RV,
the conditional probability of the configuration is
$P(C = (0,1,1,1,1) \, | F_{202})= \tau \times \tau \times \tau^2 \times \tau^2 = \tau^6$.
If IDs 203 or 209 introduce the RV, the conditional probability of the configuration is 
$P(C = (0,1,1,1,1) \, | F_{203})=P(C = (0,1,1,1,1) \, | F_{209})=0$. The likelihood function for $\tau$ is 
therefore, again, 
$P(C = (0,1,1,1,1) \, | \cup_i \! F_i) =\frac{1}{4} \tau^6$ and evaluates to $0.00390625$ for $\tau=1/2$, 
as confirmed by the code chunk below:
```{r}
config6 = c("0", "1", "1", "1", "1")
names(config6)=c("201","206","207","211", "214")
likehd(testPed2, 1/2, config6)
```

# 5. Likelihood curves

This section plots likelihood curves for the
toy pedigree created in section 1 and the 
modified `kinship2` pedigree created in section 4.2.

## 5.1 Toy pedigree

We start by considering the toy pedigree. First, 
we plot likelihood values versus a sequence of transmission 
probabilities, $\tau$, when the RV configuration for IDs 7, 8, and 9 is $C = (1,1,1)$.
```{r}
# Create a sequence of tau values
tau.seq = seq(from = 0, to = 1, by = 0.01)

#first configuration example in the toy pedigree
config1=c("1","1","1")
names(config1)=c("7","8","9")
likehd.vec = c()
for (p in tau.seq) {
  likelihood = likehd(toyPed, p, config1)
  likehd.vec = append(likehd.vec, likelihood)
}

plot(tau.seq, likehd.vec, type = "l", xlab = "transmission probabilities", ylab = "likelihood values"
     ,main = "Likelihood curve when the RV configuration is C = (1,1,1)")
```

From the above plot, we observe that, for configuration
$C=(1,1,1)$, the likelihood values increase rapidly once the transmission probability is greater than 0.6, up to a likelihood 
value of 0.5. The maximum-likelihood estimate of the
transmission probability is $\hat \tau =1$.

Next, we plot the likelihood curve for 
RV configuration $C = (1,1,0)$.
```{r}
config2=c("1","1","0")
names(config2)=c("7","8","9")
likehd.vec = c()
for (p in tau.seq) {
  likelihood = likehd(toyPed, p, config2)
  likehd.vec = append(likehd.vec, likelihood)
}

plot(tau.seq, likehd.vec, type = "l", xlab = "transmission probabilities", ylab = "likelihood values",
     main = "Likelihood curve when the RV configuration is C = (1,1,0)")
```

From the above plot, we see that for $C=(1,1,0)$ 
the likelihood curve
reaches a peak at a transmission probability of approximately 0.8. 
Likelihood values increase with the 
transmission probability when the transmission probability is less than 0.8, 
and decrease when the transmission probability is greater than 0.8.
The maximum-likelihood estimate of the 
transmission probability is thus $\hat \tau \approx 0.8$

## Modified `kinship2` pedigree

Next, we plot two likelihood curves for the 
first modified `kinship2` pedigree in section 4.2.
We start by considering the configuration $C=(1,1,1,1)$ when all four of the affected pedigree members 
with available DNA carry a single copy of the RV.
```{r}
config3 = c("1", "1", "1", "1")
names(config3)=c("206","207","211", "214")
likehd.vec = c()
for (p in tau.seq) {
  likelihood = likehd(testPed, p, config3)
  likehd.vec = append(likehd.vec, likelihood)
}

plot(tau.seq, likehd.vec, type = "l", xlab = "transmission probabilities", ylab = "likelihood values",
     main = "Likelihood curve when the RV configuration is C = (1,1,1,1)")
```

From the above plot, we see that the likelihood values increase rapidly when 
transmission probability is greater than 0.6.
The maximum likelihood estimate of the 
transmission probability in this case is
$\hat \tau =1$.

Next suppose that we observe the configuration
$C=(1,1,0,1)$. 
```{r}
config4 = c("1", "1", "0", "1")
names(config4)=c("206","207","211", "214")
likehd.vec = c()
for (p in tau.seq) {
  likelihood = likehd(testPed, p, config4)
  likehd.vec = append(likehd.vec, likelihood)
}

plot(tau.seq, likehd.vec, type = "l", xlab = "transmission probabilities", ylab = "likelihood values",
     main = "Likelihood curve when the RV configuration is C = (1,1,0,1)")

```

From the above plot, we observe that the likelihood 
curve reaches its peak value 
when the transmission probability is approximately 0.8.  Likelihood values increase with the
transmission probability when the transmission
probability is less than 0.8, and decrease 
when the transmission probability is greater than 0.8.
The maximum likelihood estimate of the transmission
probability is therefore $\hat \tau \approx 0.8$.

Finally, let us conclude by considering 
the configuration $C=(1,0,0,0)$; that is
ID 206 carries a single copy of the RV 
and IDs 207, 211 and 214 carry no copies.
```{r}
config5 = c("1", "0", "0", "0")
names(config5)=c("206","207","211", "214")
likehd.vec = c()
for (p in tau.seq) {
  likelihood = likehd(testPed, p, config5)
  likehd.vec = append(likehd.vec, likelihood)
}
plot(tau.seq, likehd.vec, type = "l", xlab = "transmission probabilities", ylab = "likelihood values",
     main = "Likelihood curve when the RV configuration is C = (1,0,0,0)")
```

From the above plot, we see that the likelihood 
curve reaches a peak value when the
transmission probability is approximately 0.4.
The maximum likelihood estimate of the 
transmission probability in this case
is $\hat \tau \approx 0.4$.

# References

[1] Højsgaard, S. . (2012). Graphical independence networks with the `gRain` package for R. Journal of Statistical Software, 46(10), 1–26.

[2] Sinnwell JP, Therneau TM, Schaid DJ. The `kinship2` R package for pedigree data. Hum Hered. 2014;78(2):91-3. doi: 10.1159/000363105. Epub 2014 Jul 29. PMID: 25074474; PMCID: PMC4154601.

